Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-06-24 manuscript.Rmd
judge what? closes #32

We show that the simple decision rule is just a special case of a more general conceptualization under Bayesian reasoning. 
To that end, we first introduce some tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theoretical risk" [a term borrowed from @meehlTheoreticalRisksTabular1978 but assigned to the probability of proving a hypothesis wrong, if it does not hold], which is inversely related to type I error rate.
We then outline two interpretations of how theoretical risk is impacted by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently by limiting type I error.
The second interpretation is our main contribution and demonstrates the broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes after preregistration.
Following this interpretation, the theoretical risk is not necessarily directly maximized by preregistration, but rather the uncertainty in judging the theoretical risk is minimized.

To arrive at this interpretation, we rely on three arguments.
The first is that theoretical risk is vital for judging theories.
The second argument is that the theoretical risk for a given study is generally uncertain.
The third and last argument is that this uncertainty is reduced by applying preregistration.
We conclude that because preregistration decreases uncertainty about the theoretical risk, which in turn increases our expectation to gain evidence for or against a theory, preregistration is warranted for any study.

@@

We already have noted that there is not much to do about $P(H)$ (and hence its counter probability $P(¬H)$), and that it is common sense to increase $P(E|H)$.
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ is theoretical risk.

@@

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set by itself is well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology, even without statistical issues like alpha inflation [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence to develop a more precise (and therefore risky) theory, e.g. by specifying which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape etc.
We will later show that preregistration increases the belief one can stake in the further specified theory, though it remains low till substantiated by testing it again.
The point, however, is that we want to show that preregistration increases the expected epistemic value without regard to the theory being tested.

@@

Preregistration is often discussed in relation to statistical methods, so let us note some relations between $P(E|¬H)$ and statistical methods.
If you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, then the nill-hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate can hence be more sure of their results, if significant.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ (and hence its inverse, theorethical risk) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

>>>>

We show that the simple decision rule is just a special case of a more general conceptualization under Bayesian reasoning. 
To that end, we first introduce some tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theoretical risk" [a term borrowed from @meehlTheoreticalRisksTabular1978 but assigned to the probability of proving a hypothesis wrong, if it does not hold], which is inversely related to type I error rate.

We then outline two interpretations of how theoretical risk is impacted by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently by limiting type I error.
The second interpretation is our main contribution and demonstrates the broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes after preregistration.
Following this interpretation, the theoretical risk is not necessarily directly maximized by preregistration, but rather the uncertainty in judging the theoretical risk is minimized.

To arrive at this interpretation, we rely on three arguments.
The first is that theoretical risk is vital for judging evidential support for theories.
The second argument is that the theoretical risk for a given study is generally uncertain.
The third and last argument is that this uncertainty is reduced by applying preregistration.
We conclude that because preregistration decreases uncertainty about the theoretical risk, which in turn increases our expectation to gain evidence for or against a theory, preregistration is warranted for any study.

@@

We already have noted that there is not much to do about $P(H)$ (and hence its counter probability $P(¬H)$), and that it is common sense to increase $P(E|H)$.
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ is theoretical risk.

@@

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set by itself is well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology, even without statistical issues like alpha inflation [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence to develop a more precise (and therefore risky) theory, e.g. by specifying which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape etc.
We will later show that preregistration increases the belief one can stake in the further specified theory, though it remains low until substantiated by testing it again.
The point, however, is that we want to show that preregistration increases the expected epistemic value independent of the theory being tested.

@@

Preregistration is often discussed in relation to statistical methods, so let us note some relations between $P(E|¬H)$ and statistical methods.
If you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, then the nill-hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate can hence be more sure of their results, if significant.
In other words, the probability that a given significant finding reflects a true effect is larger if researchers lower their type I error rates.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ (and hence its inverse, theorethical risk) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.
